---
title: "Group exercise 1"
author: "Pernille Kjeilen Fauskanger"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

\newcommand{\EE}[2]{\mathrm{E}#1[ #2 #1]}
\newcommand{\fasit}[1]{\underline{\underline{#1}}}
## Task 1 -- Proof of 2.2 a

We are to prove that
\begin{equation}
  \EE{\big}{u(w - X)} \leq \EE{\big}{u(w - Y)} \Longleftrightarrow \EE{\big}{b + a \cdot u(w - X)} \leq \EE{\big}{b + a \cdot u(w - Y)},
\end{equation}
for $a \in \mathbb{R}^+$ and $b \in \mathbb{R}$. Here, $\mathbb{R}^+ = (0, \infty)$. We begin to prove '$\Longrightarrow$':
\begin{align}
  &\EE{\big}{u(w - X)} \leq \EE{\big}{u(w - Y)} \overset{a > 0}{\Longrightarrow} a \cdot \EE{\big}{u(w - X)} \leq a \cdot \EE{\big}{u(w - Y)} \\
  &\Longrightarrow \EE{\big}{b} + a \cdot \EE{\big}{u(w - X)} \leq \EE{\big}{b} + a \cdot \EE{\big}{u(w - Y)} \nonumber \\
  &\Longrightarrow \EE{\big}{b + a \cdot u(w - X)} \leq \EE{\big}{b + a \cdot u(w - Y)} \square \nonumber
\end{align}
We now prove '$\Longleftarrow$':
\begin{align}
  &\EE{\big}{b + a \cdot u(w - X)} \leq \EE{\big}{b + a \cdot u(w - Y)} \overset{\mathrm{lin \; op}}{\Longrightarrow} \EE{\big}{b} + \EE{\big}{a \cdot u(w - X)} \leq \EE{\big}{b} + \EE{\big}{a \cdot u(w - Y)} \\
  &\Longrightarrow \EE{\big}{a \cdot u(w - X)} \leq \EE{\big}{a \cdot u(w - Y)} \nonumber \Longrightarrow a \cdot \EE{\big}{u(w - X)} \leq a \cdot \EE{\big}{u(w - Y)} \nonumber \\
  &\overset{a>0}{\Longrightarrow} \EE{\big}{u(w - X)} \leq \EE{\big}{u(w - Y)} \square \nonumber
\end{align}
We have now proved that
\begin{equation}
  \EE{\big}{u(w - X)} \leq \EE{\big}{u(w - Y)} \Longleftrightarrow \EE{\big}{b + a \cdot u(w - X)} \leq \EE{\big}{b + a \cdot u(w - Y)}.
\end{equation}

Let us define
\begin{equation}
  v(x) = \frac{u(x) - u(0)}{u'(0)}
\end{equation}
Assuming that $u(0) = 0$, $u(1) = 1$ and $u'(0) > 0$, we can see that $v$ is a linear transformation of $u$ entailing that $v$ is equivalent to $u$ based on the proof above. Defining $v$ in this way imply that $v(0) = 0$, and $v'(0) = 1$, of course, assuming that $u'(0) \ne 0$, which is the case for $u'(0) > 0$.


## Task 2 -- Exercise 1.2.4

Let $u(x) = \sqrt{x}, \; x \geq 0 \Rightarrow u(w - Z) = \sqrt{w - Z}, \; w \geq Z$. This person will prefer $X$ to $Y$ if the expected utility evaluated at $X$ is larger than at $Y$. So, we need to show that
\begin{equation}
  \EE{\big}{u(X)} > \EE{\big}{u(Y)} \Rightarrow \EE{\big}{\sqrt{X}} > \EE{\big}{\sqrt{Y}} \Rightarrow \sum P(X = x) \cdot \sqrt{x} > \sum P(X = y) \cdot \sqrt{y}
\end{equation}
This is straight forward as we know the PMFs for $X$ and $Y$:
\begin{align}
  &\sum P(X = x) \cdot \sqrt{x} > \sum P(X = y) \cdot \sqrt{y} \Rightarrow \frac{1}{2}\Big[20 + 30\Big] > \frac{1}{5}\Big[30 + 80\Big] \nonumber \\
  &\Rightarrow 25 \geq 22 \Rightarrow \text{prefer X to Y} \nonumber
\end{align}
He is risk-averse because $u''(x) < 0$. A risk-averse person that now have determined that he want to go for $X$ rather than $Y$, would decline for values of $w$ where facing $X$ is better than facing $\EE{\big}{X}$;
\begin{align}
  &25 = \EE{\big}{u(X)} > u\Big(\EE{\big}{X}\Big) = \sqrt{w} \Rightarrow w < 625 \nonumber
\end{align}
For example, a linear utility function would results in this person preferring $Y$ over $X$.

## Task 3 -- Exercise 1.2.8

Nothing about current wealth are stated in the text, so we assume that $w = 0$ for simplicity. Winning the first game yields a price of $1$. Then the utility function evaluated at his current wealth is then $u(1)$ because his new wealth is $1$. Assuming that the the chance of winning the game $0.5$, we can see that the expected utility playing a second round is $\EE{\big}{u(2)}$. The choice of yes, signify that he is a risk-lover:
\begin{equation}
  \EE{\big}{u(2)} > u(1) \Rightarrow \frac{1}{2}(u(0) + u(2)) > 1 \Rightarrow u(2) > 2
\end{equation}
If he plays a third round, his expected utility is $\EE{\big}{u(4)}$. Saying yes to this after a huddle, as he does, entails that he is more risk-neutral all of the sudden:
\begin{equation}
  \EE{\big}{u(4)} = u(2) \Rightarrow \frac{1}{2}(u(0) + u(4)) = u(2) \Rightarrow u(4) = 2u(2) > 4
\end{equation}
If he plays a fourth round, his expected utility is $\EE{\big}{u(8)}$. Saying no to this, as he does, entails that he now is risk-averse:
\begin{equation}
  \EE{\big}{u(8)} < u(4) \Rightarrow \frac{1}{2}(u(0) + u(8)) = u(4) \Rightarrow 8 > 2u(4) > u(8) 
\end{equation}

# Task 4 -- Exercise 1.3.6

We use the exponential utility function with $\alpha = 0.001$. Recall that the exponential utility function $u$ is defined by a persons wealth $w$, in the following manner:
\begin{equation}
  u(w) = -\alpha \exp(-\alpha w)
\end{equation}
The premium, $P$, may be defined by the equilibrium principle, which is defined by this relationship with the risk, $R$:
\begin{equation}
  u(w) = -\alpha \exp(-\alpha w) = \EE{\big}{u(w + P - R)} = \EE{\big}{-\alpha \exp(-\alpha(w + P - R))} \nonumber
\end{equation}
The utility functions evaluated at the wealth $w$ cancels, which simplifies the computation
\begin{align}
  1 = \EE{\big}{\exp(-\alpha(P - R))} &\Rightarrow \exp(\alpha P) = \EE{\big}{\exp(\alpha R)} \Rightarrow \alpha P = \log \EE{\big}{\exp(\alpha R)} \nonumber \\
  &\Rightarrow P = \frac{1}{\alpha} \log(M_R(\alpha))
\end{align}
We see that the premium $P$ is a function of $\alpha$ and will therefore be denoted by $P(\alpha)$. For the given risks $X$, and $Y$ (that are normally distributed, and therefore the MGFs are well defined) and $\alpha = 0.001$, we have that
\begin{align}
  \log(M_X(0.001)) &= \mu_X \cdot 0.001 + \frac{\sigma_X ^ 2 \alpha ^ 2}{2} = 400 \cdot 0.001 + \frac{1}{2} \cdot 25000 \cdot 0.001^2 = `r 400 * 0.001 + 0.5 * 25e3 * 0.001^2` \nonumber \\
  \log(M_Y(0.001)) &= \mu_Y \cdot 0.001 + \frac{\sigma_Y ^ 2 \alpha ^ 2}{2} = 420 \cdot 0.001 + \frac{1}{2} \cdot 20000 \cdot 0.001^2 = `r 420 * 0.001 + 0.5 * 20e3 * 0.001^2` \nonumber
\end{align}
Thus, the premiums for risks $X$ and $Y$ are calculated to be
\begin{align}
  P_{X}(0.001) &= \fasit{`r (400 * 0.001 + 0.5 * 25e3 * 0.001^2) / 0.001`} \\
  P_{Y}(0.001) &= \fasit{`r (420 * 0.001 + 0.5 * 20e3 * 0.001^2) / 0.001`}
\end{align}
The premium for the risk $Y$ is larger. To see which $\alpha$ satisfying that the premium of $Y$ are higher than that for $X$, we solve the following inequality for $\alpha$:
\begin{align}
  &P_{Y}(\alpha) > P_{X}(\alpha) \Rightarrow \frac{4.2 \cdot 10 ^ 2 \cdot \alpha + 0.5 \cdot 2 \cdot 10^4 \alpha^2}{\alpha} > \frac{4 \cdot 10 ^ 2 \cdot \alpha + 0.5 \cdot 2.5 \cdot 10^4 \alpha^2}{\alpha} \nonumber \\
  &\Rightarrow 4.2 \cdot 10 ^ 2 + \alpha \cdot 10^4 > 4.0 \cdot 10 ^ 2 + 1.25 \cdot \alpha \cdot 10^4 \nonumber \\
  &\Rightarrow 20 > 2.5 \cdot 10^3 \alpha \Rightarrow \underline{\underline{\alpha < `r 20 / 2500`}}
\end{align}

# Task 5 -- Exercise 1.3.8

We have that $P(\alpha) \geq 1250$. We can solve the inequality to say something about $\alpha$. As we showed in the previous task, the minimum premium the insurer are willing to agree on, is $P^-$, which was defined based on $\alpha$ and:
\begin{equation}
  P^-(\alpha) = \frac{1}{\alpha} \Big[ \mu_R \cdot \alpha + \frac{1}{2}\sigma_R^2 \alpha ^ 2 \Big]
\end{equation}
So we are interested in solving
\begin{equation}
  1250 \geq \frac{1}{\alpha} \Big[\mu_R \cdot \alpha + \frac{1}{2}\sigma_R^2 \alpha ^ 2\Big]
\end{equation}
Here $\mu_R = 10^3$ and $\sigma_R^2 = 10^4$. So we have everything we need to obtain $\alpha$ values solving this inequality:
\begin{align}
  &1250 \geq \frac{1}{\alpha} \Big[10^3 \cdot \alpha + \frac{1}{2} 10^4 \alpha ^ 2\Big] = 10^3 + 5 \cdot 10^3 \cdot \alpha \Rightarrow 50 \geq 10^3 \cdot \alpha \nonumber \\
  &\frac{50}{10^3} \geq \alpha \Rightarrow \alpha \leq \fasit{`r 50 / 1000`}
\end{align}

So what can we say about the risk aversion coefficient when it is restricted to $\alpha \in (0, `r 50 / 1000`]$? See figure \@ref(fig:risk-aversion-alpha-T5-1), to see the relationship between wealth $w$ and the exponential utility $u$.

```{r risk-aversion-alpha-T5-1, fig.width = 4, fig.height = 3, echo = FALSE, warning = FALSE, error = FALSE, fig.cap = "The exponential utility function for the alpha values representing the risk-aversion of the insurer."}

library(ggplot2)
library(Rcpp)
library(data.table)

cppFunction(code = "NumericVector exponential_utility(NumericVector w, float alpha = 0.05){
            int n = w.size();
            NumericVector eu(n);
            for(int i = 0; i < n; ++i){
              float part_1 = exp(w[i] * alpha);
              float part_2 = (-1) * alpha;
              eu[i] = part_2 * part_1;
            }
            return eu;
            }")

wealths <- seq(from = 0, to = 250, length.out = 501)
alphas <- as.list(c(0.001, 0.005, 0.01, 0.015, 0.02, 0.025, 0.03, 0.04, 0.05))
plotting_data <- lapply(X = alphas,
                        FUN = function(x) list(w = wealths, eu = exponential_utility(w = wealths, alpha = x)))
names(plotting_data) <- paste("alpha = ", unlist(alphas))
plotting_data <- rbindlist(l = plotting_data, idcol = "alpha")

ggplot(data = plotting_data, mapping = aes(x = w, y = eu * (-1), color = alpha)) + geom_line() +
  scale_x_continuous(name = "w", na.value = 0, trans = "log10") +
  scale_y_continuous(name = "-u(w)", na.value = 0, trans = "log2") + theme_bw() +
  theme(axis.text = element_blank(),
        title = element_text(face = "bold"),
        legend.background = element_rect(fill = "white", color = "black"),
        legend.key = element_rect(fill = "white", color = "black"))
  
```

What is the dimension of $\alpha$ given that the dimension of $R$ is $\mathrm{money}$? We define $\alpha$ given the distribution of $X$ by
\begin{equation}
  \alpha = (P_R(\alpha) - \mu_R) \cdot \frac{2}{\sigma_R^2}
\end{equation}
Here $\mathrm{dim}(P_R(\alpha)) = \mathrm{dim}(\mu_R) = \mathrm{dim}(\sigma_R) = 1$. The dimension of $\sigma_R^2$ is then $2$. Thus the dimension of $\alpha$ is
\begin{equation}
  \mathrm{dim}(\alpha) = \mathrm{money} \cdot \frac{1}{\mathrm{money}^2} = \fasit{\mathrm{money}^{-1}}
\end{equation}
